---
title: "Understanding the Equation of a Line"
subtitle: "$Y = \\beta X + \\epsilon$"
format:
  revealjs:
    theme: dark
    transition: slide
    highlight-style: github
    slide-number: true
    chalkboard: false
    css: styles.css
execute:
  echo: false
---

## The Full Equation {auto-animate=true}

<br>

::: {.r-fit-text}
$$Y = \beta X + \epsilon$$
:::

---

## Breaking It Down {auto-animate=true}

::: {.r-stack}

::: {.fragment .fade-in-then-semi-out}
$$\color{#4FC3F7}{Y} = \beta X + \epsilon$$

::: {.callout-note icon=false}
### 👁️ $Y$ — The Outcome
$Y$ is what we **observe directly** in the world.  
It's real, measurable, and varies across individuals or cases.  
Think of it as the **data we actually collect**.
:::
:::

::: {.fragment .fade-in-then-semi-out}
$$Y = \color{#FF8A65}{\beta} X + \epsilon$$

::: {.callout-warning icon=false}
### 🔒 $\beta$ — The Parameters (Unobserved, Indirect)
$\beta$ is **never directly seen** — we can only *estimate* it from data.  
It represents the **true relationship** between $X$ and $Y$ in the population.  
We infer $\beta$ indirectly through the data we collect.
:::
:::

::: {.fragment .fade-in-then-semi-out}
$$Y = \beta X + \color{#CE93D8}{\epsilon}$$

::: {.callout-important icon=false}
### 🌫️ $\epsilon$ — The Error (Unobserved Uncertainty)
$\epsilon$ captures everything we **didn't measure or can't explain**.  
It is unobserved — we never know it directly.  
It represents the **irreducible randomness** in the world.
:::
:::

:::

---

## $Y$ — Observed but Variable {.smaller}

$$\color{#4FC3F7}{Y} = \beta X + \epsilon$$

<br>

::: columns
::: {.column width="55%"}
::: {.fragment}
**$Y$ is observable.**  
We *see* it, *measure* it, *collect* it.

Examples:
- Income of a person
- Blood pressure of a patient  
- Test score of a student
:::

::: {.fragment}
<br>
**But $Y$ is not fixed.**  
It varies — sometimes wildly — across observations.  
That variation is *why we need a model*.
:::
:::

::: {.column width="45%"}
::: {.fragment}
```{r}
#| fig-height: 4
set.seed(42)
x <- rnorm(60, mean = 5, sd = 1.5)
y <- 2 + 1.5 * x + rnorm(60, sd = 2)
plot(x, y,
     pch = 19, col = "#4FC3F7",
     xlab = "X", ylab = "Y",
     main = "Observed Y values",
     bg = "transparent",
     col.axis = "white",
     col.lab = "white",
     col.main = "white",
     fg = "white")
```
:::
:::
:::

---

## $Y$ Has a Mean {.smaller}

$$\color{#4FC3F7}{E[Y]} = \beta X$$

<br>

::: {.fragment}
The **expected value** (mean) of $Y$ given $X$ is a *smooth function* of $X$.  
This is the signal we're trying to recover.
:::

<br>

::: {.fragment}
```{r}
#| fig-height: 3.2
set.seed(42)
x <- rnorm(60, mean = 5, sd = 1.5)
y <- 2 + 1.5 * x + rnorm(60, sd = 2)
plot(x, y, pch = 19, col = "#4FC3F7",
     xlab = "X", ylab = "Y",
     main = "E[Y|X] — the conditional mean of Y",
     col.axis = "white", col.lab = "white",
     col.main = "white", fg = "white")
abline(lm(y ~ x), col = "#FF8A65", lwd = 3)
legend("topleft", legend = "E[Y|X] = βX", col = "#FF8A65",
       lwd = 3, bty = "n", text.col = "white")
```
:::

---

## $\beta$ — Unobserved, Indirect {.smaller}

$$Y = \color{#FF8A65}{\beta} X + \epsilon$$

<br>

::: columns
::: {.column width="60%"}
::: {.fragment}
**$\beta$ is the true population parameter.**  
It exists in theory — but we **never observe it directly**.
:::

::: {.fragment}
<br>
We only ever have **estimates** $\hat{\beta}$ computed from data.

$$\hat{\beta} = (X'X)^{-1}X'Y$$

These estimates are **indirect** — they depend entirely on what we observed in $Y$ and $X$.
:::

::: {.fragment}
<br>
::: {.callout-warning icon=false}
### The Estimation Gap
True $\beta$ ≠ Estimated $\hat{\beta}$  
The difference is sampling uncertainty.
:::
:::
:::

::: {.column width="40%"}
::: {.fragment}
```{r}
#| fig-height: 4.5
set.seed(123)
betas <- replicate(500, {
  x <- rnorm(30, 5, 1.5)
  y <- 2 + 1.5 * x + rnorm(30, sd = 2)
  coef(lm(y ~ x))[2]
})
hist(betas, col = "#FF8A65", border = "white",
     main = "Distribution of β̂\nacross 500 samples",
     xlab = "Estimated β",
     col.axis = "white", col.lab = "white",
     col.main = "white", fg = "white")
abline(v = 1.5, col = "white", lwd = 2, lty = 2)
legend("topright", legend = "True β = 1.5",
       col = "white", lwd = 2, lty = 2,
       bty = "n", text.col = "white")
```
:::
:::
:::

---

## $\epsilon$ — Unobserved Uncertainty {.smaller}

$$Y = \beta X + \color{#CE93D8}{\epsilon}$$

<br>

::: {.fragment}
**$\epsilon$ is the error term — permanently unobserved.**  

It captures everything that $X$ doesn't explain:
- Measurement error  
- Omitted variables  
- Inherent randomness in the world
:::

::: {.fragment}
<br>

::: columns
::: {.column width="50%"}
We **assume** $\epsilon \sim N(0, \sigma^2)$

- Mean zero: no systematic bias  
- Finite variance: uncertainty is bounded  
- Independent of $X$: errors aren't predictable
:::
::: {.column width="50%"}
```{r}
#| fig-height: 3
set.seed(42)
x <- rnorm(60, 5, 1.5)
y <- 2 + 1.5 * x + rnorm(60, sd = 2)
resid <- residuals(lm(y ~ x))
plot(x, resid, pch = 19, col = "#CE93D8",
     xlab = "X", ylab = "Residuals (proxy for ε)",
     main = "Residuals ≈ ε (but not equal!)",
     col.axis = "white", col.lab = "white",
     col.main = "white", fg = "white")
abline(h = 0, col = "white", lwd = 2, lty = 2)
```
:::
:::
:::

---

## Putting It All Together {auto-animate=true}

<br>

::: {.r-fit-text}
$$\color{#4FC3F7}{Y} = \color{#FF8A65}{\beta} X + \color{#CE93D8}{\epsilon}$$
:::

<br>

| Term | Color | Observed? | Nature |
|------|-------|-----------|--------|
| $Y$ | [Blue]{style="color:#4FC3F7"} | ✅ Yes — directly | Outcome; has a mean |
| $\beta$ | [Orange]{style="color:#FF8A65"} | ❌ No — indirectly | Population parameter; estimated |
| $\epsilon$ | [Purple]{style="color:#CE93D8"} | ❌ No — never | Random uncertainty |

---

## The Key Insight {background-color="#1a1a2e"}

<br><br>

::: {.r-fit-text .fragment}
We observe **$Y$**
:::

::: {.r-fit-text .fragment}
We estimate **$\beta$**
:::

::: {.r-fit-text .fragment}
We assume **$\epsilon$**
:::

::: {.fragment}
<br>

> *Statistics is the art of learning about the unobserved ($\beta$, $\epsilon$) from the observed ($Y$).*
:::
